Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2015-01-14 paper.tex
Minor fixes addressing comments from Thomas.

%Why is it hard? (E.g., why do naive approaches fail?)
Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
%The evaluation of threshold queries of derived fields requires a stencil kernel computation at each point on the grid in order to compute the field to be thresholded.
%The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
%In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
%will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
It is impractical to materialize all possible derived fields and store them alongside the raw data
due to the large size of the datasets and the limits of available storage.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step took over 20 hours. It would take months to iterate over the entire dataset. 
This highlighted the need for providing the capability through an integrated approach, which performs the evaluation server-side.

@@

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and applica\-tion-level
information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached. 
We will demonstrate that an integrated approach, which computes the derived fields on-demand in a data-parallel manner, performs the evaluation over 
an entire time-step in a few minutes. Storing the query results in an application-aware semantic cache further reduces the running times to several 
seconds.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
large multidimensional array datasets. 
Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
Each data point appears in $N$ kernels of computation, in which $N$ is the size of the kernel.
The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
the network and to achieve scalability.
Techniques that target the traditional supercomputing environments do not translate directly to the 
distributed database setting of an analysis cluster environment.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. 
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. 
%Each computation is distributed across the nodes of the cluster and
%is executed in parallel by several processes within a database node by creating multiple connections to each node. 
We extend the database management system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation in a limited version of a semantic cache. 
This leads to query performance improvement of over an order of magnitude.

@@

\section{JHTDB}\label{jhtdb}
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built into the database through user-defined functions or stored procedures.

@@

Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
cached in an application-aware cache, which is a limited version of a semantic cache. 
The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.

@@

The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available SSD disk space, not memory.
Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
space and database overhead is $\sim$40MB. 
Therefore for a dataset containing 
1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
require $\sim$5TB (15TB for vector fields).

@@

We observe nearly a two times speedup when going from a single process per node to two processes per node (Fig. \ref{fig:scaleup}). 
The speedup diminishes to 1.4 times when going to 4 processes and little speedup is 
observed with 8 processes per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region are requested by multiple processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
compared to running with a single process.

@@

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast, our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
examines the vector or scalar array data at all neighboring locations, which
are within the kernel of the computation. The functions used to derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

>>>>

%Why is it hard? (E.g., why do naive approaches fail?)
Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
%The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
%The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
%In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
%will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
It is impractical to materialize all possible derived fields and store them alongside the raw data
due to the large size of the datasets and the limits of available storage.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step took over 20 hours. It would take months to iterate over the entire dataset. 
This highlighted the need for providing the capability through an integrated approach, which performs the evaluation server-side.

@@

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
sufficient for some applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached. 
We will demonstrate that an integrated approach, which computes the derived fields on-demand in a data-parallel manner, performs the evaluation over 
an entire time-step in a few minutes. Storing the query results in an application-aware semantic cache further reduces the running times to several 
seconds.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. 
The field variables have to be evaluated \emph{on-demand} from the array data stored in the database cluster. 
The evaluations are data-intensive as they perform kernel computations on extremely large multidimensional array datasets. 
A kernel computation computes the value at a grid location using the data points at a set of neighboring locations.
Kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
the network and to achieve scalability.
Techniques that target the traditional supercomputing environments do not translate directly to the 
distributed database setting of an analysis cluster environment.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. 
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. 
%Each computation is distributed across the nodes of the cluster and
%is executed in parallel by several processes within a database node by creating multiple connections to each node. 
We extend the database management system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache query results along with query metadata and subsequent queries are evaluated against the cache.
This leads to query performance improvement of over an order of magnitude.

@@

\section{Johns Hopkins Turbulence\\ Databases}\label{jhtdb}
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built into the database through user-defined functions or stored procedures.

@@

Threshold queries of derived fields submitted to the JHTDB are evaluated using a data-parallel execution strategy and the query results are 
cached in an application-aware semantic cache. 
In addition to query results, the cache stores their semantic descriptions and query metadata and parameters used to obtain them.
The evaluation strategy for queries that do not hit the cache is driven by the spatial partitioning of the data across the nodes of the cluster.

@@

The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available SSD disk space, not memory.
Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
space and database overhead is $\sim$40MB. 
Therefore for a dataset containing 
1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
results for threshold queries of a derived field over the entire dataset. The currently available SSD disk space per node is $\sim$200GB, which will
be sufficient to maintain the threshold results for nearly five derived fields over the entire dataset. 
In contrast, computing and materializing a scalar derived field for the entire dataset would
require $\sim$5TB (15TB for vector fields).

@@

We observe nearly a two times speedup when going from a single process per node to two processes per node (Fig. \ref{fig:scaleup}). 
The speedup diminishes to 1.4 times when going to 4 processes and little speedup is 
observed with 8 processes per node.
While the computation time scales with increased process count, the time to perform I/O does not as the data on each node reside in the same 
database table and on the same set of disks. Additionally, I/O redundancy increases as the process count increases as data along the boundary 
of each region are requested by multiple processes.
SQL Server already utilizes parallelism to perform the I/O even when data are retrieved utilizing a single query.
Finally, the experiments were run on the live production database nodes, which were also servicing other user queries in addition to operating system
and other SQL Server processes. Nevertheless, running with 4 processes per node is nearly 2.6 times faster when
compared to running with a single process.

@@

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast, our approach performs a kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
examines the vector or scalar array data at all neighboring locations, which
are within the kernel of the computation. The functions used to derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.
